---
title: "Lab#2"
author: "Ifeoma Okonye"
date: "2024-01-30"
output: html_document
---


#Introduction

To begin this lab session, I installed the *gapminder* data. I then called the library and displayed the first six rows of my dataset to survey the dataset and see the different variables in the data.

```{r}
#install.packages("gapminder")
library(gapminder)
data("gapminder")
head(gapminder, 6)
```

## Question 1

Next, I re-ordered the data in reverse alphabetical order of countries and viewed the first 15 entries.

```{r}
country_reverse <- gapminder[order(gapminder$country, decreasing = TRUE), ]
head(country_reverse, 15)
```


## Question 2

I then calculated and displayed the mean, median and standard deviation of the Life  Life expectancy, Population, and GDPperCapita in the gapminder dataset.

```{r}
#Statistic for Life Expectancy
mean(gapminder$lifeExp)
median(gapminder$lifeExp)
sd(gapminder$lifeExp)

#Statistic for Population
mean(gapminder$pop)
median(gapminder$pop)
sd(gapminder$pop)

#Statistic for GDP per Capita
mean(gapminder$gdpPercap)
median(gapminder$gdpPercap)
sd(gapminder$gdpPercap)
```


## Question 3

Additionally, I calculated the mean, median and standard deviation of the Life  Life expectancy, Population, and GDPperCapita for each country and continent.

```{r}
#Calculating the mean, median and standard deviation for life expectancy, population and GDP per capita by country.

mean_stat <- aggregate(gapminder[ , c("lifeExp", "pop", "gdpPercap")], 
                       by = gapminder["country"], FUN = mean, na.action = rm)
head(mean_stat, 5)

median_stat <- aggregate(gapminder[ , c("lifeExp", "pop", "gdpPercap")], 
                         by = gapminder["country"], FUN = median, na.action = rm)
head(median_stat, 5)


sd_stat <- aggregate(cbind(lifeExp, pop, gdpPercap) ~ country, 
                     data = gapminder, FUN = sd, na.rm = TRUE)
head(sd_stat, 5)


#Calculating the mean, median and standard deviation for life expentancy, population and gdp per capita by continent.

mean_stat1 <- aggregate(gapminder[ , c("lifeExp", "pop", "gdpPercap")], 
                        by = gapminder["continent"], FUN = mean, na.action = rm)
head(mean_stat1)

median_stat1 <- aggregate(gapminder[ , c("lifeExp", "pop", "gdpPercap")], 
                          by = gapminder["continent"], FUN = median, na.action = rm)
head(median_stat1)

sd_stat1 <- aggregate(gapminder[ , c("lifeExp", "pop", "gdpPercap")], 
                      by = gapminder["continent"], FUN = function(x) sd(x, na.rm = T)) #alternative method
head(sd_stat1)
```


## Question 4

I created a scatter plot of the global average Life expectancy over time and displayed the first 5 records for the dataset used to build the plot. I went ahead to Customize the display of the plot to make it easier to understand.

```{r}
avg_lifeExp <- aggregate(gapminder[ , "lifeExp"], by = gapminder["year"], 
                         FUN = mean, na.action = rm)
head(avg_lifeExp, 5)

time <- avg_lifeExp$year
lifeExp <- avg_lifeExp$lifeExp
plot(time, lifeExp,  panel.first = grid(8, 8),
     main = "Life Expectancy over Time", col = "blue", bg = "red", pch = 21, cex = 2)
points(time, lifeExp, col = "black", pch = 16, cex = 0.5)
```


## Question 5

I then pulled in the **Air Quality** dataset from the readily available R datasets and examined the structure. I looked at the class types, header information and the dimensions.

```{r}
data(airquality)
a_qal <- airquality 

class(a_qal)
sapply(a_qal, class)

head(a_qal, 10)

ncol(a_qal)
nrow(a_qal)
dim(a_qal)
```


## Question 6 and 7

I went ahead to calculate the general summary for the entire dataset. and then I calculated summary statistics for groups/subsets of the dataset.

```{r}
summary(a_qal, na.rm = TRUE)


summary(a_qal$Ozone, na.rm = TRUE)
summary(a_qal$Temp, na.rm = TRUE)
summary(a_qal$Solar.R, na.rm = TRUE)
summary(a_qal$Wind, na.rm = TRUE)
summary(a_qal$Month, na.rm = TRUE)
summary(a_qal$Day, na.rm = TRUE)
```


## Question 8

Finally, I created 2 plots, one showing ungrouped data and the other showing grouped data. I gave each plot a title, descriptive x and y axis labels and customized a few graphic option on the plots.

```{r}
# Ungrouped Plot
hist(a_qal$Temp, main = "Frequency of Temperature Variation", breaks = 12, 
     col = "skyblue", xlab = "Temperature")

# Grouped Plot
plot(a_qal$Ozone, a_qal$Solar.R, 
     xlab = "Ozone", ylab = "Solar Radiation", 
     main = "Plot showing correlation between Ozone and Solar Radiation",
     pch = 21, col = "black", bg = as.factor(a_qal$Month), cex = 2)

?hist
```

